---
title: "Potential Aquaculture"
subtitle: "West Coast Suitable Growing Conditions"
author: "Yos Ramirez"
format: html
editor: visual
execute: 
  echo: true
  warning: false
  message: false
---

# Prioritizing potential aquaculture

## Prepare Data

```{r}
suppressMessages({
# Load required libraries
library(raster)     # For raster operations 
library(sp)         # For spatial operations 
library(sf)         # For shapefile handling 
library(ggplot2)    # For data visualization 
library(dplyr)      # For data manipulation 
library(terra)      # For raster and vector manipulation (alternative to 'raster')
library(tmap)       # For thematic mapping (visualization)
})
```

```{r}
# One line to read, reproject, and convert to SpatialPolygonsDataFrame
west_coast_eez_sp <- as_Spatial(st_transform(st_read("data/wc_regions_clean.shp"), crs = "+proj=utm +zone=10 +datum=WGS84"))
# Load bathymetry raster (RasterLayer object)
bathymetry_raster <- raster("data/depth.tif")
# Load SST rasters (RasterStack object)
sst_rasters <- stack(list.files(path = "data/", pattern = "sst.*.tif", full.names = TRUE))
```

```{r, cache=TRUE}
# Set the random seed for reproducibility
set.seed(123)  # Set a fixed seed value for reproducibility
# Check the number of layers and names in the SST stack
nlayers(sst_rasters)
names(sst_rasters)
```

```{r}
# Check the CRS of each spatial object
crs_eez <- proj4string(west_coast_eez_sp)  # CRS of the EEZ shapefile
crs_bathymetry <- projection(bathymetry_raster)  # CRS of the bathymetry raster
crs_sst <- projection(sst_rasters)  # CRS of the SST raster stack
  
# Print individual CRS values to debug
print(paste("EEZ CRS: ", crs_eez))
print(paste("Bathymetry CRS: ", crs_bathymetry))
print(paste("SST CRS: ", crs_sst))

```

```{r}
# Convert CRS to the target CRS (if it's not already in UTM Zone 10N)
if (crs_bathymetry != "+proj=utm +zone=10 +datum=WGS84") {
  bathymetry_raster <- projectRaster(bathymetry_raster, crs = "+proj=utm +zone=10 +datum=WGS84")
  print("Reprojected bathymetry raster to match UTM Zone 10N CRS.")
} else {
  print("CRS of bathymetry raster already matches UTM Zone 10N.")
}

if (crs_sst != "+proj=utm +zone=10 +datum=WGS84") {
  sst_rasters <- projectRaster(sst_rasters, crs = "+proj=utm +zone=10 +datum=WGS84")
  print("Reprojected SST raster stack to match UTM Zone 10N CRS.")
} else {
  print("CRS of SST raster stack already matches UTM Zone 10N.")
}

```

# (Building workflow for function)

## Process Data 

```{r}
# Parameters for oysters
suitable_sst_min <- 11  # Minimum suitable SST for oysters in Celsius
suitable_sst_max <- 30  # Maximum suitable SST for oysters in Celsius
suitable_depth_min <- 0  # Minimum suitable depth in meters
suitable_depth_max <- 70  # Maximum suitable depth in meters
species_name <- "Oysters"

```

```{r}
# Calculate mean SST over the desired period (e.g., 2008-2012)
sst_mean <- mean(sst_rasters[[names(sst_rasters)[grepl(paste(2008:2012, collapse = "|"), names(sst_rasters))]]], na.rm = TRUE)


# Convert SST from Kelvin to Celsius
sst_celsius <- sst_mean - 273.15
crs_celsius <- projection(sst_celsius)

# Reproject SST raster stack to match EEZ CRS (if needed)
if (crs_celsius != "+proj=utm +zone=10 +datum=WGS84") {
  sst_celsius <- projectRaster(sst_celsius, crs = "+proj=utm +zone=10 +datum=WGS84")
  print("Reprojected SST raster stack to match UTM Zone 10N CRS.")
} else {
  print("CRS of SST raster stack already matches UTM Zone 10N.")
}

# Resample bathymetry raster and crop to extent of sst_celsius
bathymetry_raster <- resample(crop(bathymetry_raster, extent(sst_celsius)), sst_celsius, method = "ngb")

```

## Find Suitable Locations

```{r}
# Set suitable and unsuitable areas in bathymetry
bathymetry_suitable <- calc(bathymetry_raster, fun = function(x) {
  ifelse(is.na(x) |x <= suitable_depth_min | x >= suitable_depth_max, 0, 1)
})

# Plot the bathymetry suitability map
plot(bathymetry_suitable, main = "Suitable Depth (Bathymetry)", xlab = "Eastings (m)", ylab = "Northings (m)", legend=FALSE)
```

```{r}
# Set suitable and unsuitable areas in Sea Surface Temperature
sst_suitable <- calc(sst_celsius, fun = function(x) {
  ifelse(is.na(x) | x <= suitable_sst_min | x >= suitable_sst_max, 0, 1)
})

# Plot the SST suitability map
plot(sst_suitable, main = "Suitable Sea Surface Temperature", xlab = "Eastings (m)", ylab = "Northings (m)",legend=FALSE)
```

```{r}
# Combine the suitability rasters using overlay (multiplication for both suitable conditions), both need to be 1 for suitability
suitable_locations <- overlay(bathymetry_suitable, sst_suitable, fun = function(x, y) {
  x * y  
})

# Plot the resulting suitability map
plot(suitable_locations,main = "Suitable Locations", xlab = "Eastings (m)", ylab = "Northings (m)",legend=FALSE)
```

## Determine the Most Suitable EEZ

```{r}
# Rasterize the EEZ shapefile to match the resolution of the suitable locations raster
eez_raster <- rasterize(west_coast_eez_sp, suitable_locations, field = "rgn_id", fun = mean)

# Plot the EEZ raster
plot(eez_raster, main = "Exclusive Economic Zones", xlab = "Eastings (m)", ylab = "Northings (m)")
```

```{r, cache=TRUE}
# Set the random seed for reproducibility
set.seed(123)  # Set a fixed seed value for reproducibility
# Calculate the total suitable area by EEZ region using zonal(), calculate cell area
suitable_area_by_eez <- zonal(suitable_locations, eez_raster, fun = "sum", na.rm = TRUE)
cell_area <- res(suitable_locations)[1] * res(suitable_locations)[2]

# Multiply by cell area to get the total area (assuming suitable locations have values 1)
total_area_by_eez <- suitable_area_by_eez[,2] * cell_area

# View the results
print(total_area_by_eez)
```

```{r}
# Merge the total area data with the EEZ shapefile to prepare for plotting
west_coast_eez_sp$Total_Area <- total_area_by_eez
# View Columns of EEZ shapefile
head(west_coast_eez_sp@data)
```

## Map

```{r}
# Plotting the West Coast EEZ (Exclusive Economic Zone) shapefile
tm_shape(west_coast_eez_sp) +
  tm_borders() +  # Adding borders for the EEZ boundaries
  tm_fill(col = "Total_Area",  # Color the polygons based on the 'Total_Area' attribute
          palette = "viridis",  # Color palette for the fill (Viridis color scheme)
          title = "Total Area (m²)") +  # Title for the color legend
  tm_text("Total_Area",  # Text labels showing 'Total_Area'
          size = 0.7,  # Text size
          col = "black",  # Text color
          fontface = "bold",  # Bold font for the labels
          bg.col = "white",  # Background color for text labels
          bg.alpha = 0.7) +  # Background transparency
  tm_layout(main.title = paste(species_name, "Suitability by EEZ"),  # Title with species name
            legend.outside = TRUE,  # Place the legend outside the map
            legend.position = c("left", "bottom"),  # Position of the legend
            legend.width = 2,  # Width of the legend
            legend.title.size = 1.2,  # Font size for the legend title
            legend.text.size = 0.8,  # Font size for the legend text
            legend.title.fontface = "bold",  # Bold font for the legend title
            legend.text.fontface = "bold") +  # Bold font for the legend text
  tm_scale_bar(position = c("left", "bottom")) +  
  tm_compass(position = c("right", "top"))  # Position of the compass

```
### Reflection

The map above for the oyster species shows the exclusive economic zones with the area suitability, specific to its temperature and depth preferences. The zone with the most area in m2 for oyster aquaculture is Washington. Southern California has the second largest suitable area for oyster aquaculture and Central California follows. Interestingly Central California has the least suitable area for oyster aquaculture, noticeable in the "Suitable Locations" mask.
The area is calculated from the x and y pixels in m2 and multiplied with 0 or 1 to give the final numbers. The set.seed allows for reproducibility and helps produce the same output each time. I kept everything in Large RasterLayer along with the West Coast EEZ Spatial Polygons and transformed crs to UTM 10 for meters in the west coast. Overall the approach to the workflow was to minimize the code but have clear output visualizations to understand each step.

# Function for any species

```{r, cache=TRUE}
# General function to determine suitable locations for marine aquaculture
calculate_suitable_eez_area <- function(sst_rasters, bathymetry_raster, eez_shapefile, 
                                        suitable_sst_min, suitable_sst_max, 
                                        suitable_depth_min, suitable_depth_max, 
                                        species_name, seed = 123) {
  
  # Set the random seed for reproducibility
  set.seed(seed)
  # 1. Calculate mean SST over the desired period (e.g., 2008-2012)
  sst_mean <- mean(sst_rasters[[names(sst_rasters)[grepl(paste(2008:2012, collapse = "|"), names(sst_rasters))]]], na.rm = TRUE)
  
  # 2. Convert SST from Kelvin to Celsius
  sst_celsius <- sst_mean - 273.15
  
  # 3. Reproject SST raster stack to match EEZ CRS (if needed)
  crs_celsius <- projection(sst_celsius)
  if (crs_celsius != "+proj=utm +zone=10 +datum=WGS84") {
    sst_celsius <- projectRaster(sst_celsius, crs = "+proj=utm +zone=10 +datum=WGS84")
    print("Reprojected SST raster stack to match UTM Zone 10N CRS.")
  } else {
    print("CRS of SST raster stack already matches UTM Zone 10N.")
  }
  
  # 4. Resample bathymetry to match SST raster
  bathymetry_raster <- resample(crop(bathymetry_raster, extent(sst_celsius)), sst_celsius, method = "ngb")
  
  # 5. Calculate bathymetry suitability
  bathymetry_suitable <- calc(bathymetry_raster, fun = function(x) {
    ifelse(is.na(x) | x <= suitable_depth_min | x >= suitable_depth_max, 0, 1)
  })
  
  # 6. Calculate SST suitability
  sst_suitable <- calc(sst_celsius, fun = function(x) {
    ifelse(is.na(x) | x <= suitable_sst_min | x >= suitable_sst_max, 0, 1)
  })
  
  # 7. Combine suitability maps (overlay)
  suitable_locations <- overlay(bathymetry_suitable, sst_suitable, fun = function(x, y) {
    x * y  # Both need to be 1 for suitability
  })
  
  # 8. Rasterize EEZ shapefile to match suitable locations raster
  eez_raster <- rasterize(eez_shapefile, suitable_locations, field = "rgn_id", fun = mean)
  
  # 9. Calculate total suitable area by EEZ region
  suitable_area_by_eez <- zonal(suitable_locations, eez_raster, fun = "sum", na.rm = TRUE)
  cell_area <- res(suitable_locations)[1] * res(suitable_locations)[2]
  total_area_by_eez <- suitable_area_by_eez[, 2] * cell_area
  
  # 10. Merge the total area data with the EEZ shapefile
  eez_shapefile$Total_Area <- total_area_by_eez
  
  # 11. Plot the results using tmap
  map <- tm_shape(eez_shapefile) + 
    tm_borders() + 
    tm_fill(col = "Total_Area", 
            palette = "viridis", 
            title = "Total Area (m²)") +
    tm_text("Total_Area", 
            size = 0.7, 
            col = "black", 
            fontface = "bold",
            bg.col = "white", 
            bg.alpha = 0.7) +
    tm_layout(main.title = paste(species_name, "Suitability by EEZ"), 
              legend.outside = TRUE,
              legend.position = c("left", "bottom"),         
              legend.width = 2,                             
              legend.title.size = 1.2,                      
              legend.text.size = 0.8,                     
              legend.title.fontface = "bold",               
              legend.text.fontface = "bold")+
  tm_scale_bar(position = c("left", "bottom")) +  
  tm_compass(position = c("right", "top"))
  
  return(map)
}
```

### Squid

```{r}
# Define the species name for the map title or further analysis
species_name <- "Squid"  # Setting the species name for easy reference in plots or output
suitable_sst_min <- 8  # Minimum sea surface temperature (in °C) for squid suitability
suitable_sst_max <- 13  # Maximum sea surface temperature (in °C) for squid suitability
suitable_depth_min <- 20  # Minimum suitable depth (in meters) for squid
suitable_depth_max <- 180  # Maximum suitable depth (in meters) for squid

map_squid <- calculate_suitable_eez_area(sst_rasters,        # Sea Surface Temperature raster data
                                          bathymetry_raster,  # Bathymetry (depth) raster data
                                          west_coast_eez_sp,  # EEZ spatial data
                                          suitable_sst_min,   # Minimum SST for suitability
                                          suitable_sst_max,   # Maximum SST for suitability
                                          suitable_depth_min, # Minimum depth for suitability
                                          suitable_depth_max, # Maximum depth for suitability
                                          species_name)       # Name of the species 

# Print the resulting map or summary of the suitable areas for the species
print(map_squid) 

```

### Reflection

The map above for the squid species shows the exclusive economic zones with the area suitability, specific to its temperature and depth preferences. Squid is suitable in more areas than oyster, ranging up to 1200 million m2 in Washington. The total area for California zones are much closer to the oyster areas in the same zones but still differ. It would be interesting to see the "Suitable Locations" mask for squid. I assume the area would be further from the coast because of the colder temperature and extent of greater depth.
The function can be used with different max depth, min depth, max temperature, min temperature, and species name. I kept only the necessary code for that portion after evaluating what was necessary from the workflow. I set a seed for reproducibility and cache = TRUE to ensure the same output each time it is ran and rendered.

#### References

Hijmans, R. J. (2023). raster: Geographic Data Analysis and Modeling (R package version 3.6-23). Retrieved from https://CRAN.R-project.org/package=raster.
Bivand, R. S., Pebesma, E. J., & Gómez-Rubio, V. (2013). Applied Spatial Data Analysis with R (2nd ed.). Springer.
Pebesma, E. J. (2018). Simple features for R: Standardized support for spatial vector data. R Journal, 10(1), 439-446. https://doi.org/10.32614/RJ-2018-009.
Wickham, H. (2016). ggplot2: Elegant Graphics for Data Analysis. Springer. https://doi.org/10.1007/978-3-319-24277-4.
Wickham, H., & Francois, R. (2020). dplyr: A Grammar of Data Manipulation (R package version 1.0.4). Retrieved from https://CRAN.R-project.org/package=dplyr.
Hijmans, R. J. (2022). terra: Spatial Data Analysis (R package version 1.6-32). Retrieved from https://CRAN.R-project.org/package=terra.
Tennekes, M. (2018). tmap: Thematic Maps in R. Journal of Statistical Software, 84(6), 1-39. https://doi.org/10.18637/jss.v084.i06.
